Reduced plasticity and variance in physiological rates for terrestrial ectotherm populations under climate change

### 2023-12-14 ms.qmd
Think I have a rough first draft done

## **Abstract**

@@

<!-- 
  1. Phenotypic plasticity and evolutionary adatation to novel environments
  2. Capacity for plasticity and adaptation differ between aqautic and terrestrial ectotherms
  3. Climate change is expected to impact on the capacity for plasticity and adaptation
  4. We test whether climate variability and predictability explain a populations acclimation capacity and affect changes in trait variance.
-->

@@

Periods of past climatic change have had disproportionate impacts on some ecosystems over others leading to debates over which ecosystems will be most vulnerable to contemporary climate change. Studies have highlighted species occupying terrestrial ecosystems as being particularly vulnerable given their weak acclimation abilities and greater probability of experiencing thermal extremes that overwhelm physiological homeostasis [@Seebacher2015-wi; @Gunderson2015-hz; @Pinsky2019-fn; @hoffmann2013upper]. Despite marine and freshwater ecosystems appearing to have greater physiological acclimation capacity [e.g., @Seebacher2015-wi; @Pottier2022], it is unclear if the magnitude of physiological adjustment is sufficient. Low oxygen availability has been suggested as a major factor influencing the vulnerability of aquatic ecosystems whereas remaining close to thermal limits and greater water loss is expected to be a stronger constraint on physiological processes in terrestrial ectotherms [@Verberk2016]. Given that terrestrial ectotherms are expected to be closer to their thermal limits increases in temperature may have a stronger impact on variation in physiological rates within populations compared to aqautic ectotherms. Reduced variability may result from variances being eroded by strong selection, an inability to adjust physiological rates, or a combination of both. Changes in physiological rate variance may therefore have important cascading effects on energy transfer and productivity for populations across these different ecosystems that are as of yet poorly measured and modelled.

@@

Monthly average temperature data were extracted from the ERA5 climate model, available from the Copernicus climate data store [@Hersbach2020-hx]. For each population and species in the dataset we extracted a 30-year period (1950-2022) of either surface air temperature at 2 meters (0.01 resolution) for both terrestrial and freshwater taxa, or sea surface temperature for the marine taxa (at 0.25 resolution) using the *ncdf4* R package [vers. `r utils::packageVersion("ncdf4")`, @Pierce2021-lc]. We chose a 2-meter resolution because we believed that it more likely to reflects the micro-thermal environment experienced by terrestrial and freshwater ectotherms at those locations. For terrestrial species we also collected soil temperature as for many species this maybe more representative of microhabitat choice compared to air temperature. We fit models using both air and soil temperature and found that the results were qualitatively similar. As such, we only present results for air temperature (See Supplementary Materials).

@@

#### *Multi-level Meta-analysis (MLMA) Models*
We first fit multi-level meta-analysis (MLMA) models (i.e., intercept only models) for both $lnRR_{Q_{10}}$ and $lnCVR_{Q_{10}}$, that included study, species, and phylogeny as random effects to account for non-independence. We also included trait as a random effect to account for trait variation within the data. Our MLMA models allowed us to partition the variation in $lnRR_{Q_{10}}$ and $lnCVR_{Q_{10}}$ among these key sources while accounting for total sampling variance in each. This allowed us to calculate total heterogeneity [i.e., $I_{total}^2$; *sensu* @Nakagawa2012-oc; @Noble2022-ty] along with various $I^2$ metrics describing the proportion of variance explained by each random effect level [@Nakagawa2012-oc]. We also present 95% prediction intervals which describe the expected distribution of effects from future studies [@Nakagawa2021-ls; @Noble2022-ty].

@@

### *The opportunity for selection is reduced in terrestrial ectotherms*

The opportunity for selection, as captured by $lnCVR_{Q_{10}}$, showed a decrease across all habitat types but was especially pronouced in terrestrial and marine ectotherms. Overall, there was a `r mean(t_cvr_mean_ef)`% (95% CI: `r quantile(t_cvr_mean_ef, 0.025)[1]` to `r quantile(t_cvr_mean_ef, 0.975)[1]`, $p_{MCMC}$ = `r pmcmc(t_cvr_mean)`) reduction in variation in physiological rates for terrestrial ectotherms and a `r mean(m_cvr_mean_ef)`% (95% CI: `r quantile(m_cvr_mean_ef, 0.025)[1]` to `r quantile(m_cvr_mean_ef, 0.975)[1]`, $p_{MCMC}$ = `r pmcmc(m_cvr_mean)`) reduction in variation for marine ectotherms when temperatures increase by 10$^\circ$C. In contrast, freshwater ectotherms exhibited a much smaller reduction in physiological rate variance at high temperatures (`r mean(f_cvr_mean_ef)`%, 95% CI: `r quantile(f_cvr_mean_ef, 0.025)[1]` to `r quantile(f_cvr_mean_ef, 0.975)[1]`, $p_{MCMC}$ = `r pmcmc(f_cvr_mean)`).

@@

Thermal variability (i.e., $CV$) and predictability a population experienced in its past did not explain acclimation capacity or changes in opportunity for selection (i.e., the change in physiological variability) among terrestrial, marine or freshwater populations (@fig-4; @fig-5).

@@

### Global change in opportunity for selection

@@

Acute and acclimation responses for wild ectotherms were much less common than studies done on captive populations ($N_{species}$ = `r length(unique(all_pred$spp))`, from `r length(unique(all_pred$uniques))` populations). Globally, there was a clear bias towards spiecies in the Northern Hemiphere (@fig-fig6 A-C). Projected changes in the opportunity for selection were highly variable across the globe, with some regions showing a decrease in the opportunity for selection, while others showing an increase (@fig-fig6 D). Out of the `r length(unique(all_pred2$match))` populations variance was predicted to decrease in `r (sum(decreases$n, na.rm = TRUE) / length(decreases$n[complete.cases(decreases$n)]))*100`% of the locations.

Predictions of current global changes in the opportunity for selection were generally conservative with our model explaining ~ 50% of the variation in the observed data ($R^2$ = `r round(mean(r2_2), 2)`, 95% CI: `r round(quantile(r2_2, 0.025)[1], 2)` to `r round(quantile(r2_2, 0.975)[1], 2)`). Across habitat types climate change is predicted to result in a only a `r abs(lapply(mean_perc_habitat, mean)$f)`% change in variance for freshwater systems (95% CI: `r lapply(mean_perc_habitat, function(x) quantile(x, c(0.025, 0.975)))$f[1]` to `r lapply(mean_perc_habitat, function(x) quantile(x, c(0.025, 0.975)))$f[2]`%, $p_{MCMC}$ = `r pmcmc(mean_perc_habitat$f)`), whereas we expect a `r abs(lapply(mean_perc_habitat, mean)$m)`% reduction in variance for marine systems (95% CI: `r lapply(mean_perc_habitat, function(x) quantile(x, c(0.025, 0.975)))$m[1]` to `r lapply(mean_perc_habitat, function(x) quantile(x, c(0.025, 0.975)))$m[2]`%, $p_{MCMC}$ = `r pmcmc(mean_perc_habitat$m)`) and a `r abs(lapply(mean_perc_habitat, mean)$t)`% reduction in variance for terrestrial systems (95% CI: `r lapply(mean_perc_habitat, function(x) quantile(x, c(0.025, 0.975)))$t[1]` to `r lapply(mean_perc_habitat, function(x) quantile(x, c(0.025, 0.975)))$t[2]`%, $p_{MCMC}$ = `r p_value(pmcmc(mean_perc_habitat$t))`) under a RCP8.5 climate scenerio.

@@

Understanding acclimation capacity and how variation in physiological rates change across populations and species is important for predicting the ecological and evolutionary consequences of climate change [@Chevin2017; @Chevin2010; @Bolnick2011]. Here we show that variance in physiological rates is expected to decrease for ectotherms as temperatures increase, but particularly so for terrestrial ectotherms (~28%). Decreases in variance of physiological rates is larger than the expected benefits from acclimation in marine and terrestrial ectotherms (marine: 9% acclimation vs. 15% reduction in variance; terrestrial: no acclimation vs 28% reduction in variance). Such effects are expected to result in a decrease in the variability in physiologcial rates for marine and terrestrial ectotherms between ~4-13% under future climate change projections. Our results provide a more complete picture of the impacts that climate warming will have on ectotherm populations, with important ecological and evolutionary implications that require further empirical and theoretical attention.

### ***Acclimation capacity and opportunity for selection across ectotherms in response to climate change***

There is a general interest in understanding the interplay between plasticity and genetic adaptation to a populations capacity to mitigate climate change effects []. We show that plastic adjustments to physiological rates are rather small (if they exist at all) compared to the general reduction variance in physiological rates. While it is unclear the extent to which such changes in variance are heritable we expect changes in variance of this magnitude to have an impact on the opportunity for selection by reducing the strength of selection or underlying genetic variance. Both are predcited to reduce the capacity of populations to evolve under climate change. One explanation for why terrestrial ectotherms show minimal acclimation capacity may be related to the fact that terrestrial ectotherms, were, on average acclimated for significantly less time than ectotherms from aquatic habitats. However, rates of acclimation have been shown to be faster for many terrestrial groups compared to aquatic organisms [e.g., amphibians and reptiles have higher rates of acclimation than fishes; See @Einum2023-lz]. Faster rates of acclimation would therefore result in the opposite patterns to those we observed -- in other words, terrestrial species would be more likely to exhibit lower $lnRR_{Q_{10_{acclimation}}}$ compared to $lnRR_{Q_{10_{acute}}}$ when controlling for acclimation time. 
assuming a fraction of this variation is heritable and/or it weakens selection

@@

Theoretical models predict that plasticity should evolve in populatons experiencing greater environmental variability (spatial or temporal), particularly if oscillations are predictable over time making cues more reliable [@Chevin2017; @Lande2009; @Chevin2010; @Nunney2016; @Murren2015; @hendry2016key; @Reed2010]. Higher spatial and temporal heterogenetiy in terrestrial ecosystems [@steele2019comparison] suggests that plasticity should be more likely evolve in terrestrial environments, however, we did not find evidence to this effect.  The lack of relationship between acclimation responses and environmental variability and predictability could be due to the challenges in estimating realistic microthermal enviroments experienced by taxa or because lower predictability of thermal environments over time make cues unreliable preventing plastic responses from evolving easily [@Leung2020; @hendry2016key]. Randomly flutuating environments have been suggested to select for reduced plasticity, particularly if plastic responses are costly and this has been demonstrated by a number of empirical studies [@Leung2020; @rescan2022experimental; @leung2023phenotypic]. For example, using seed beetles (*Callosobruchus maculatus*), @hallsson2012selection showed that experimentally evolving populations under randomly flutuating thermal conditions did not lead to the expected increase in plasticity, but reduced plasticity. @Leung2020 also showed reduced plasticity in morphological traits when environments experienced were less predictable by experimentally evolving algae populations (*Dunaliella salina*) [see also @leung2023phenotypic]. The fact that aquatic ectotherms did show a capacity to acclimate suggest that environmental predcitability, which is likely generally higher in aquatic environments, may be more important than environmental variability in driving the evolution of plasticity.

### ***Ecological consequences of reduced variation in physiological Rates***
While we found that variance in physiological rates decreases substantially at high temperatures, especially pronouced in terrestrial ecosystems, variance reductions in metabolism and performance were key traits most strongly impacted. Changes in variability in physiological rates, particularly traits governing energy demand, could have important consequences on the flow of energy within and between populations, communities and ecosystems [@hendry2016key]. More variable populations, genetically and/or phenotypically, are predicted to be associated with broader niches, reduced intraspecific competition, increased growth rate, decreased vulnerability to environmental change and lower extinction risk [@forsman2015rethinking; @forsman2014effects; @Bolnick2011; @hendry2016key]. Maintaining intrapopulaton variability in physiological rates in a warmer world may therefore be important for population resilience to climate change. Nonetheless, few studies have explored the consequences that variation in physiological traits have for population dynamics. Those studies that have have shown that such variation can have important implications. For example, @kooijman1989population integrated individual variation in dynamic energy budget (DEB) models developed for *Daphnia magna*. They showed that greater between-individual variation in DEB parameters led to less extreme population occilations compared to scenerios with no variation [@kooijman1989population]. The differences in variance change across major habitat types suggest that terrestrial ecotherms will likely face an onslaught of ecological and evolutionary challenges which will mean greater rates of extintion and reduced productivity – predictions already reinforced by a number of studies [@hoffmann2013upper; @Gunderson2015; @Pottier2022]. To date, however, an appreciation for the wide-reaching impacts that changes in variance may have on populations and communities remains limited. The ecological consequences of changes in physiological rate variance across aquatic and terrestrial ecosystems remain unclear. What we desperately need now are the development of theoretical models that better integrate intrapopulation variation into mechanistic physiological models or ones that better capture stochasticity such as individual based models. Our meta-analytic findings can help us better position ourselves to now start asking questions about the ecological and evolutionary consequences of changes in variance in physiological rates.

### ***Limitations and future directions***

An enhanced knowledge of how variation in physiological rates vary across populations and species, and the degree to which they can be adjusted in response to the environment, may increase our ability to make more informed predictions about ecological and evolutonary dynamics of natural populations [@forsman2015rethinking]. Variance in physiological rates could be better incoperated in physiological and ecological models to provide more nuanced, and possibly more realistic, predictions about the impacts of climate change on natural populations. Our meta-analysis mean that such models can be that are better aligned with empirical findings. Nonethless, many fascinating questions remain unanswered that will require greater focus on the consequences of change sin variance (rather than just the mean): How does a reduction in physiological rate varaince change the flow of energy across tropic levels within communities? What are the biochemical, cellular and physiological mechanisms that underly reduced variance in physiological rates? What fraction of the change in variance is environmental versus genetic?

>>>>

## **Abstract**
 will write this later...should take me 20 minutes

@@

@@

Periods of past climatic change have had disproportionate impacts on some ecosystems over others leading to debates over which ecosystems will be most vulnerable to contemporary climate change. Studies have highlighted species occupying terrestrial ecosystems as being particularly vulnerable given their weak acclimation abilities and greater probability of experiencing thermal extremes that overwhelm physiological homeostasis [@Seebacher2015-wi; @Gunderson2015-hz; @Pinsky2019-fn; @hoffmann2013upper]. Despite marine and freshwater ecosystems appearing to have greater physiological acclimation capacity [e.g., @Seebacher2015-wi; @Pottier2022], it is unclear if the magnitude of physiological adjustment is sufficient, particularly in combination with potential changes other factors such as genetic and phenotypic variability. Low oxygen availability has been suggested as a major factor influencing the vulnerability of aquatic ecosystems whereas remaining close to thermal limits and greater water loss is expected to be a stronger constraint on physiological processes in terrestrial ectotherms [@Verberk2016]. Given that terrestrial ectotherms are expected to be closer to their thermal limits increases in temperature may have a stronger impact on variation in physiological rates within populations compared to aqautic ectotherms. Reduced variability may result from variances being eroded by strong selection, an inability to adjust physiological rates, or a combination of both. Changes in physiological rate variance may therefore have important cascading effects on energy transfer and productivity for populations across these different ecosystems that are as of yet poorly measured and modelled.

@@

Monthly average temperature data were extracted from the ERA5 climate model, available from the Copernicus climate data store [@Hersbach2020-hx]. For each population and species in the dataset we extracted a 30-year period (1950-2022) of either surface air temperature at 2 meters (0.01 resolution) for both terrestrial and freshwater taxa, or sea surface temperature for the marine taxa (at 0.25 resolution) using the *ncdf4* R package [vers. `r utils::packageVersion("ncdf4")`, @Pierce2021-lc]. We chose a 2-meter resolution because we believed that it more likely to reflects the micro-thermal environment experienced by terrestrial and freshwater ectotherms at those locations. For terrestrial species we also collected soil temperature as for many species this maybe more representative of microhabitat choice compared to air temperature. We fit models using both air and soil temperature and found that the results were qualitatively similar. As such, we only present results for air temperature.

@@

#### *Multi-level Meta-analysis (MLMA) Models*
We first fit multi-level meta-analysis (MLMA) models (i.e., intercept only models) for both $lnRR_{Q_{10}}$ and $lnCVR_{Q_{10}}$, that included study, species, and phylogeny as random effects to account for non-independence. We also included trait as a random effect to account for trait variation within the data. Our MLMA models allowed us to partition the variation in $lnRR_{Q_{10}}$ and $lnCVR_{Q_{10}}$ among these key sources while accounting for total sampling variance in each. This allowed us to calculate the proportion of total heterogeneity [i.e., $I_{total}^2$; *sensu* @Nakagawa2012-oc; @Noble2022-ty] along with various $I^2$ metrics describing the proportion of variance explained by each random effect level [@Nakagawa2012-oc]. We also present 95% prediction intervals which describe the expected distribution of effects from future studies [@Nakagawa2021-ls; @Noble2022-ty].

@@

### *Reduced variation in physiological rates in terrestrial ectotherms*

Variance in physiological rates, as captured by $lnCVR_{Q_{10}}$, showed a decrease across all habitat types but was especially pronouced in terrestrial and marine ectotherms. Overall, there was a `r mean(t_cvr_mean_ef)`% (95% CI: `r quantile(t_cvr_mean_ef, 0.025)[1]` to `r quantile(t_cvr_mean_ef, 0.975)[1]`, $p_{MCMC}$ = `r pmcmc(t_cvr_mean)`) reduction in variation in physiological rates for terrestrial ectotherms and a `r mean(m_cvr_mean_ef)`% (95% CI: `r quantile(m_cvr_mean_ef, 0.025)[1]` to `r quantile(m_cvr_mean_ef, 0.975)[1]`, $p_{MCMC}$ = `r pmcmc(m_cvr_mean)`) reduction in variation for marine ectotherms when temperatures increase by 10$^\circ$C. In contrast, freshwater ectotherms exhibited a much smaller reduction in physiological rate variance at high temperatures (`r mean(f_cvr_mean_ef)`%, 95% CI: `r quantile(f_cvr_mean_ef, 0.025)[1]` to `r quantile(f_cvr_mean_ef, 0.975)[1]`, $p_{MCMC}$ = `r pmcmc(f_cvr_mean)`).

@@

Thermal variability (i.e., $CV$) and predictability a population experienced in its past did not explain acclimation capacity or changes in the change in physiological variability among terrestrial, marine or freshwater populations (@fig-4; @fig-5).

@@

### Changes in physiological rate variance under climate change

@@

Acute and acclimation responses for wild ectotherms were much less common than studies done on captive populations ($N_{species}$ = `r length(unique(all_pred$spp))`, from `r length(unique(all_pred$uniques))` populations). Globally, there was a clear bias towards spiecies in the Northern Hemiphere (@fig-fig6 A-C). Projected changes in physiological rate variance were highly variable across the globe, with some regions showing a decrease in physiological rate variance, while others showing an increase (@fig-fig6 D). However, out of the `r length(unique(all_pred2$match))` populations variance was predicted to decrease in `r (sum(decreases$n, na.rm = TRUE) / length(decreases$n[complete.cases(decreases$n)]))*100`% of the locations.

Predictions of current global changes in physiological rate variance were generally conservative with our model explaining ~ 50% of the variation in the observed data ($R^2$ = `r round(mean(r2_2), 2)`, 95% CI: `r round(quantile(r2_2, 0.025)[1], 2)` to `r round(quantile(r2_2, 0.975)[1], 2)`). Across habitat types climate change is predicted to result in a only a `r abs(lapply(mean_perc_habitat, mean)$f)`% change in variance for freshwater systems (95% CI: `r lapply(mean_perc_habitat, function(x) quantile(x, c(0.025, 0.975)))$f[1]` to `r lapply(mean_perc_habitat, function(x) quantile(x, c(0.025, 0.975)))$f[2]`%, $p_{MCMC}$ = `r pmcmc(mean_perc_habitat$f)`), whereas we expect a `r abs(lapply(mean_perc_habitat, mean)$m)`% reduction in variance for marine systems (95% CI: `r lapply(mean_perc_habitat, function(x) quantile(x, c(0.025, 0.975)))$m[1]` to `r lapply(mean_perc_habitat, function(x) quantile(x, c(0.025, 0.975)))$m[2]`%, $p_{MCMC}$ = `r pmcmc(mean_perc_habitat$m)`) and a `r abs(lapply(mean_perc_habitat, mean)$t)`% reduction in variance for terrestrial systems (95% CI: `r lapply(mean_perc_habitat, function(x) quantile(x, c(0.025, 0.975)))$t[1]` to `r lapply(mean_perc_habitat, function(x) quantile(x, c(0.025, 0.975)))$t[2]`%, $p_{MCMC}$ = `r p_value(pmcmc(mean_perc_habitat$t))`) under a RCP8.5 climate scenerio.

@@

Understanding acclimation capacity and how variation in physiological rates change across populations and species is important for predicting the ecological and evolutionary consequences of climate change [@Chevin2017; @Chevin2010; @Bolnick2011]. Here we show that variance in physiological rates is expected to decrease for ectotherms as temperatures increase, but particularly so for terrestrial ectotherms (~28%). Decreases in variance of physiological rates is larger than the expected benefits from acclimation in marine and terrestrial ectotherms (*marine*: 9% acclimation vs. 15% reduction in variance; *terrestrial*: no acclimation vs 28% reduction in variance). Such effects are expected to result in a decrease in the variability in physiologcial rates for marine and terrestrial ectotherms between ~4-13% under future climate change projections. Our results provide a more complete picture of the impacts that climate warming will have on ectotherm populations, with important ecological and evolutionary implications that require further empirical and theoretical attention. We discuss our findings in the context of the current literature and highlight key areas for future research.

### ***Potential evolutionary consequences for reduced plasticity and variance in physiologcial rates across ectotherms***

There is a general interest in understanding the interplay and relative importance of plasticity and genetic adaptation to a population's capacity to mitigate climate change effects [@urban2023and; @Chevin2010; @hoffmann2011climate; @merila2014climate]. Phenotypic plasticty is expected to allow populations to persist in the face of climate extremes by better allowing phenotypes to match fitness optima, however, maladaptive responses can occur [@noble2019plastic; @radersma2020plasticity] and there may be limits / costs to plasticty that may prevent it from evolving [@Dewitt1998; but see @Murren2015]. While there is substantial variation in effects across species, we show that plastic adjustments to physiological rates are, on average, rather small compared to the general reduction variance in physiological rates. Variance reductions are particularly pronouced for terrestrial ectotherms and there were no major differences between life-stages. While it is unclear the extent to which such changes in variance are heritable, we expect changes in variance of this magnitude to have an impact on the 'opportunity for selection' by reducing the strength of selection and/or impacting underlying genetic variance exposed to selection. Both are predicted to impact the capacity of populations to evolve under climate change [@hoffmann2011climate; @urban2023and].

One explanation for why terrestrial ectotherms show minimal acclimation capacity may be related to the fact that terrestrial ectotherms, were, on average acclimated for significantly less time than ectotherms from aquatic habitats. However, rates of acclimation have been shown to be faster for many terrestrial groups compared to aquatic organisms [e.g., amphibians and reptiles have higher rates of acclimation than fishes; See @Einum2023-lz]. Faster rates of acclimation would therefore result in the opposite patterns to those that we observed -- in other words, terrestrial species would be more likely to exhibit lower $lnRR_{Q_{10_{acclimation}}}$ compared to $lnRR_{Q_{10_{acute}}}$ when controlling for acclimation time. The weak acclimation responses of terrestrial ectotherms suggests that evolving plasticty in physiological rates may be hard, possibly because terrestrial ectotherms are more likely to experience more variable and unpredictable thermal environments compared to aquatic ectotherms (see below). Reductions in physiological rate variance for terrestrial ectotherms may be because of selective mortality at higher temperatures reducing intrapopulation variability or populations being pushed closer to their thermal maxima (or a combination of both). Disentangling these hypotheses will require empirical studies to connect measures of survival at different temperatures with thermal physiology across indiviudals (e.g., $CT_{max}$).

### ***Plasticity and variance in physiologcial rates do not differ between life-stages***
Greater plasticity for early life-stages in development may be particularly important for dealing with the effects of climate change because early life-stages are usually particularly vulnerable periods in development [@martin2015age; @stearns1976life]. Evidence across the literature conflicts in terms of the extent to which one life-stage is more or less plastic than another [@carter2020life; @moghadam2019heat]. For example, @moghadam2019heat showed that larvae *Drosophila* were more plastic in their heat hardening responses compared to adults. In contrast, @carter2020life showed greater thermal plasticity in metabolism for adults but little in pupae of *Onthophagus taurus*. Across a diversity of taxa, we show that there were generally similar patterns between early and late life stages -- both in terms of variance changes and the capacity for plasticty. Unsurprisingly, there is much variation in the literature and our findings suggest that patterns will depend on the species and traits in question. However, very few studies measured multiple life-history stages within the same population. As such, we need more studies that explicitly compare, within populations, differences in plasticity among life stages across multiple traits. A better understanding of the microthermal enviroments experienced by different life-stages will also be important in understanding why plasticty (or lack thereof) may differ among life-stages.

@@

Theoretical models predict that plasticity should evolve in populatons experiencing greater environmental variability (spatial or temporal), particularly if oscillations are predictable over time making cues more reliable [@Chevin2017; @Lande2009; @Chevin2010; @Nunney2016; @Murren2015; @hendry2016key; @Reed2010]. Higher spatial and temporal heterogenetiy in terrestrial ecosystems [@steele2019comparison] suggests that plasticity should be more likely evolve in terrestrial environments, however, we did not find evidence to this effect.  The lack of relationship between acclimation responses and environmental variability and predictability could be due to the challenges in estimating realistic microthermal enviroments experienced by taxa or because lower predictability of thermal environments over time make cues unreliable preventing plastic responses from evolving easily [@Leung2020; @hendry2016key]. Randomly flutuating environments have been suggested to select for reduced plasticity, particularly if plastic responses are costly and this has been demonstrated by a number of empirical studies [@Leung2020; @rescan2022experimental; @leung2023phenotypic]. For example, using seed beetles (*Callosobruchus maculatus*), @hallsson2012selection showed that experimentally evolving populations under randomly flutuating thermal conditions did not lead to the expected increase in plasticity, but rather reduced plasticity. @Leung2020 also showed reduced plasticity in morphological traits when environments experienced were less predictable by experimentally evolving algae populations (*Dunaliella salina*) [see also @leung2023phenotypic]. The fact that aquatic ectotherms did show a capacity to acclimate suggest that environmental predictability, which is expected to be higher in aquatic environments, may be more important than environmental variability in driving the evolution of plasticity. However, disentangling the relative importance of environmental variability and predictability will require better distinguishing between active and passive plasticty, as the latter is expected to more likely be adaptive (@Havird2020-mj).

### ***Ecological consequences of reduced variation in physiological rates***
While we found that variance in physiological rates decreases substantially at high temperatures, being especially pronouced in terrestrial ectotherms, variance reductions in metabolism and performance were key traits most strongly impacted. Changes in variability in physiological rates, particularly traits governing energy demand, could have important consequences on the flow of energy within and between populations, communities and ecosystems [@hendry2016key]. More variable populations, genetically and/or phenotypically, are predicted to be associated with broader niches, reduced intraspecific competition, increased growth rate, decreased vulnerability to environmental change and lower extinction risk [@forsman2015rethinking; @forsman2014effects; @Bolnick2011; @hendry2016key]. Maintaining intrapopulaton variability in physiological rates in a warmer world may therefore be important for population resilience to climate change. Nonetheless, few studies have explored the consequences that variation in physiological traits have for population dynamics. Those studies that have, have shown that such variation can have important implications. For example, @kooijman1989population integrated individual variation in dynamic energy budget (DEB) models developed for *Daphnia magna*. They showed that greater between-individual variation in DEB parameters led to less extreme population occilations compared to scenerios with no variation [@kooijman1989population]. The differences in variance change across major habitat types suggest that terrestrial ecotherms will likely face an onslaught of ecological and evolutionary challenges which will mean greater rates of extintion and reduced productivity – predictions reinforced by a number of studies [@hoffmann2013upper; @Gunderson2015-hz; @Pottier2022]. To date, however, an appreciation for the wide-reaching impacts that changes in variance may have on populations and communities remains limited. The ecological consequences of changes in physiological rate variance across aquatic and terrestrial ecosystems remain unclear. What we desperately need now are the development of theoretical models that better integrate intrapopulation variation into mechanistic physiological models or ones that better capture stochasticity such as individual based models. Our meta-analytic findings can help us better position ourselves to now start asking questions about the ecological and evolutionary consequences of changes in variance in physiological rates.

### ***Conclusions and future directions***

An enhanced knowledge of how variation in physiological rates vary across populations and species, and the degree to which they can be adjusted in response to the environment, may increase our ability to make more informed predictions about ecological and evolutonary dynamics of natural populations [@forsman2015rethinking]. Variance in physiological rates could be better incorperated into physiological and ecological models to provide more nuanced, and possibly more realistic, predictions about the impacts of climate change on natural populations. While we do not yet understand the relative contribution of environmental and genetic factors to variances changes, models could better decouple how different levels of heritability with different total variance impact evolutionary and ecological predcitions. Our meta-analysis allows for realistic models to be parameterised and ensure they are better aligned with empirical findings. Nonetheless, many fascinating questions remain unanswered that will require greater focus on the consequences of changes in variance (rather than just the mean). Particularly interesting questions include: How does a reduction in physiological rate varaince change the flow of energy across tropic levels within communities? What are the biochemical, cellular and physiological mechanisms that underly reduced variance in physiological rates? Are reductions in variance in one trait associated with reductions in correlated traits or do some traits increase while others decrease? How much of the reduction in variance driven by lower levels of genetic variance? Answers to these questions will require integrative approaches that combine empirical and theoretical work across multiple levels of biological organisation.
